---
title: "QB2023_Team_Project"
author: "Erica Nadolski, Z620: Quantitative Biodiversity, Indiana University"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
geometry: margin = 2.54cm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(dplyr)

### Project site by species matrix



Ponds97 <- as.matrix(read.csv("/Users/ericanadolski/GitHub/TeamProject/Ponds97.csv", row.names=1))
# raw read data matrix

Ponds.pa <- as.matrix(read.csv("/Users/ericanadolski/GitHub/TeamProject/Ponds.pa.csv", row.names=1))
# presence/absence matrix

Ponds.rel <- as.matrix(read.csv("/Users/ericanadolski/GitHub/TeamProject/Ponds.rel.csv", row.names=1))
# relative abundance matrix

Ponds.env <- as.matrix(read.csv("/Users/ericanadolski/GitHub/TeamProject/Ponds.env.csv")) # environmental data

nrow(Ponds.env) # number of sites
ncol(Ponds.env)

### Rank abundance plot of site 1 
ponds.rac <- as.numeric(RAC(Ponds97[1,]))
length(ponds.rac)
max(ponds.rac)
min(ponds.rac)
plot.new()
pond.ranks <- as.vector(seq(1, length(ponds.rac)))
plot(pond.ranks, log(ponds.rac), type="p", axes=F,
     xlab="Rank in abundance", ylab="Abundance",
     las=1, cex.lab=1.4, cex.axis=1.25);
box();
axis(side=1, labels=T, cex.axis=1.25);
axis(side=2, las=1, cex.axis=1.25,
     labels= c(1,10,100,1000,100000), at=log(c(1,10,100,1000,10000)))


## Scatter plot of some env variables
env <- as.data.frame(Ponds.env)

env <- env %>% mutate_at(c("lat","long","Elevation","Diameter","Depth","Cal_Volume","ORP","Temp","SpC","DO","TDS","Salinity","pH","Color","chla","DOC","DON","canopy","TP"), as.numeric)

str(env)

diam <- env[1:56,"Diameter"]
max(diam)
min(diam)
mean(diam)

depth <- env[1:56,"Depth"]
max(depth)
min(depth)
mean(depth)

p <- ggplot(env, aes(x=Diameter, y=Depth, label=X))
p + geom_point(aes(colour = factor(Location))) + geom_text(hjust=0, vjust=0, size=2) + scale_x_continuous(n.breaks = 10)


```

> ***Code Analysis***: We are using a Lennon lab dataset of microbial DNA and cDNA extracted from local ponds. There are 58 pond sites, and 34059 species (OTUs) across all the sites. There is high variance in abundance and evenness across all of the sites; based on exploratory rank abundance curves, there is high abundance of a few OTUs and a long tail of low-abundance OTUs.
```

## PCA of Environmental Variables

```{r pressure, echo=FALSE}
## no missing data in subset: env[1:53,3:21]

env.pca = prcomp(env[1:53,3:21])
?prcomp()

```
Beta Diversity - Visualization

```{r}
# relative abundance matrix
Ponds.rel <- as.matrix(read.csv("/Users/ericanadolski/GitHub/TeamProject/Ponds.rel.csv", row.names=1))
#sites 54-58 are missing some environmental data

# subsetting total (DNA) and active (cDNA) libraries for each site

keep_DNA <- c("BC001-DNA","BC002-DNA","BC003-DNA","BC004-DNA","BC005-DNA","BC010-DNA","BC015-DNA","BC016-DNA","BC018-DNA","BC020-DNA","BC048-DNA","BC049-DNA","BC051-DNA","BC105-DNA","BC108-DNA","BC262-DNA","BCL01-DNA","BCL03-DNA","HNF132-DNA","HNF133-DNA","HNF134-DNA","HNF144-DNA","HNF168-DNA","HNF185-DNA","HNF187-DNA","HNF189-DNA","HNF190-DNA","HNF191-DNA","HNF216-DNA","HNF217-DNA","HNF221-DNA","HNF224-DNA","HNF225-DNA","HNF229-DNA","HNF236-DNA","HNF237-DNA","HNF242-DNA","HNF250-DNA","HNF267-DNA","HNF269-DNA","HNF279-DNA","YSF004-DNA","YSF117-DNA","YSF295-DNA","YSF296-DNA","YSF298-DNA","YSF300-DNA","YSF44-DNA","YSF45-DNA","YSF46-DNA","YSF47-DNA","YSF65-DNA","YSF66-DNA","YSF67-DNA","YSF69-DNA","YSF70-DNA","YSF71-DNA","YSF74-DNA")

keep_cDNA <- c("BC001-cDNA","BC002-cDNA","BC003-cDNA","BC004-cDNA","BC005-cDNA","BC010-cDNA","BC015-cDNA","BC016-cDNA","BC018-cDNA","BC020-cDNA","BC048-cDNA","BC049-cDNA","BC051-cDNA","BC105-cDNA","BC108-cDNA","BC262-cDNA","BCL01-cDNA","BCL03-cDNA","HNF132-cDNA","HNF133-cDNA","HNF134-cDNA","HNF144-cDNA","HNF168-cDNA","HNF185-cDNA","HNF187-cDNA","HNF189-cDNA","HNF190-cDNA","HNF191-cDNA","HNF216-cDNA","HNF217-cDNA","HNF221-cDNA","HNF224-cDNA","HNF225-cDNA","HNF229-cDNA","HNF236-cDNA","HNF237-cDNA","HNF242-cDNA","HNF250-cDNA","HNF267-cDNA","HNF269-cDNA","HNF279-cDNA","YSF004-cDNA","YSF117-cDNA","YSF295-cDNA","YSF296-cDNA","YSF298-cDNA","YSF300-cDNA","YSF44-cDNA","YSF45-cDNA","YSF46-cDNA","YSF47-cDNA","YSF65-cDNA","YSF66-cDNA","YSF67-cDNA","YSF69-cDNA","YSF70-cDNA","YSF71-cDNA","YSF74-cDNA")

# Active libraries - only 55
active <- Ponds.rel[rownames(Ponds.rel) %in% keep_cDNA, ]      # Extract rows from matrix
 
# total libraries - 58
total <- Ponds.rel[rownames(Ponds.rel) %in% keep_DNA, ]      # Extract rows from matrix
row.names(total) <- c("BC001","BC002","BC003","BC004","BC005","BC010","BC015","BC016","BC018","BC020","BC048","BC049","BC051","BC105","BC108","BC262","BCL01","BCL03","HNF132","HNF133","HNF134","HNF144","HNF168","HNF185","HNF187","HNF189","HNF190","HNF191","HNF216","HNF217","HNF221","HNF224","HNF225","HNF229","HNF236","HNF237","HNF242","HNF250","HNF267","HNF269","HNF279","YSF004","YSF117","YSF295","YSF296","YSF298","YSF300","YSF44","YSF45","YSF46","YSF47","YSF65","YSF66","YSF67","YSF69","YSF70","YSF71","YSF74")
```

```{r}
# Bray Curtis resemblance matrix
total.db <- vegdist(total, method="bray")

# Heatmap 
order.p <- rev(attr(total.db, "Labels"))

levelplot(as.matrix(total.db)[,order.p], aspect="iso", col.regions=inferno,
          xlab="Pond Site", ylab= "Pond Site", scales=list(cex=0.5), 
          main= "Bray-Curtis Distance")

# Wards cluster analysis
total.ward <- hclust(total.db, method= "ward.D2")

par(mar = c(1, 5, 2, 2) + 0.1)
plot(total.ward, main="Indiana Pond Bacteria: Ward's Clustering", ylab= "Squared Bray-Curtis Distance")

# Principal Component Analysis
total.pcoa <- cmdscale(total.db, eig=TRUE, k=3)

exvar1 <- round(total.pcoa$eig[1] / sum(total.pcoa$eig), 3) * 100
exvar2 <- round(total.pcoa$eig[2] / sum(total.pcoa$eig), 3) * 100
exvar3 <- round(total.pcoa$eig[3] / sum(total.pcoa$eig), 3) * 100
total.sum.eig <- sum(exvar1, exvar2, exvar3)

par(mar = c(5, 5, 1, 2) + 0.1)
total.pcoa <- plot(total.pcoa$points[ ,1], total.pcoa$points[ ,2], ylim = c(-0.2, 0.7),
     xlab= paste("PCoA 1 (", exvar1, "%)", sep = ""),
     ylab= paste("PCoA 2 (", exvar2, "%)", sep = ""),
     pch = 16, cex = 2.0, type = "n", cex.lab = 1.5,
     cex.axis=1.2, axes=FALSE);
axis(side = 1, labels = T, lwd.ticks = 2, cex.axis = 1.2, las = 1);
axis(side = 2, labels = T, lwd.ticks = 2, cex.axis = 1.2, las = 1);
abline(h = 0, v = 0, lty = 3);
box(lwd = 2);
points(total.pcoa$points[ ,1], total.pcoa$points[ ,2],
       pch = 19, cex = 3, bg = "gray", col = "gray");
text(total.pcoa$points[ ,1], total.pcoa$points[ ,2],
     labels = row.names(total.pcoa$points))
```


## SYNTHESIS week 4

Load the dataset from that you and your partner are using for the team project. 
Use one of the hypothesis-testing tools introduced in the beta diversity module. 
Interpret the findings of your data with respect to principles of biodiversity. 

```{r, echo=FALSE}
library(dplyr)

## Data input and setup 
Ponds.env <- as.matrix(read.csv("/Users/ericanadolski/GitHub/TeamProject/Ponds.env.csv", row.names=1)) # environmental data
Ponds.rel <- as.matrix(read.csv("/Users/ericanadolski/GitHub/TeamProject/Ponds.rel.csv", row.names=1))

# env data missing for "HNF279", 189, 190, 191, 236, 237

keep_DNA <- c("BC001-DNA","BC002-DNA","BC003-DNA","BC004-DNA","BC005-DNA","BC010-DNA","BC015-DNA","BC016-DNA","BC018-DNA","BC020-DNA","BC048-DNA","BC049-DNA","BC051-DNA","BC105-DNA","BC108-DNA","BC262-DNA","BCL01-DNA","BCL03-DNA","HNF132-DNA","HNF133-DNA","HNF134-DNA","HNF144-DNA","HNF168-DNA","HNF185-DNA","HNF187-DNA","HNF216-DNA","HNF217-DNA","HNF221-DNA","HNF224-DNA","HNF225-DNA","HNF229-DNA","HNF242-DNA","HNF250-DNA","HNF267-DNA","HNF269-DNA","YSF004-DNA","YSF117-DNA","YSF295-DNA","YSF296-DNA","YSF298-DNA","YSF300-DNA","YSF44-DNA","YSF45-DNA","YSF46-DNA","YSF47-DNA","YSF65-DNA","YSF66-DNA","YSF67-DNA","YSF69-DNA","YSF70-DNA","YSF71-DNA","YSF74-DNA") # "HNF189-DNA","HNF190-DNA","HNF191-DNA","HNF236-DNA","HNF237-DNA","HNF279-DNA")

total <- Ponds.rel[rownames(Ponds.rel) %in% keep_DNA, ]      # Extract rows from matrix
row.names(total) <- c("BC001","BC002","BC003","BC004","BC005","BC010","BC015","BC016","BC018","BC020","BC048","BC049","BC051","BC105","BC108","BC262","BCL01","BCL03","HNF132","HNF133","HNF134","HNF144","HNF168","HNF185","HNF187","HNF216","HNF217","HNF221","HNF224","HNF225","HNF229","HNF242","HNF250","HNF267","HNF269","YSF004","YSF117","YSF295","YSF296","YSF298","YSF300","YSF44","YSF45","YSF46","YSF47","YSF65","YSF66","YSF67","YSF69","YSF70","YSF71","YSF74") # "HNF189","HNF190","HNF191","HNF236","HNF237","HNF279")


env <- apply(Ponds.env[,1:20], 2, function(x) as.numeric(as.character(x))) # get data to numeric type
row.names(env) <- c("BC001","BC002","BC003","BC004","BC005","BC010","BC015","BC016","BC018","BC020","BC048","BC049","BC051","BC105","BC108","BC262","BCL01","BCL03","HNF132","HNF133","HNF134","HNF144","HNF168","HNF185","HNF187","HNF189","HNF190","HNF191","HNF216","HNF217","HNF221","HNF224","HNF225","HNF229","HNF236","HNF237","HNF242","HNF250","HNF267","HNF269","HNF279","YSF004","YSF117","YSF295","YSF296","YSF298","YSF300","YSF44","YSF45","YSF46","YSF47","YSF65","YSF66","YSF67","YSF69","YSF70","YSF71","YSF74")
env <- env[1:53,-1] # remove sites with NA; tried na.action=na.exclude(), na.roughfix but didn't work
env <- env[-34,]
```

### i.  Mantel Test
In the R code chunk below, do the following:   

1. create distance matrices for both fish communities and environmental factors, and
2. use a Mantel test to test the hypothesis that fish assemblages are correlated with stream environmental variables.

```{r}
# define matrices
OTU.dist <- vegdist(total, method="bray")
pond.env.dist <- vegdist(env, method="euclid")

# mantel test
mantel(OTU.dist,pond.env.dist)

```

### Constrained Ordination of Indiana Ponds data

1. create an environmental matrix of the pond data using forward and reverse selection of variables
2. conduct a redundancy analysis on the bacterial assemblages
3. use a permutation test to determine the significance of the constrained analysis,
4. use a permutation test to determine the correlation of each environmental factor on the constrained axes,
5. calculate the explained variation on the first and second constrained axes,
6. plot the constrained ordination results including labeled points for each site, and
7. add vectors that demonstrate the influence of each environmental factor the constrained ordination.   

```{r}
# Constrained Ordination
pond.dbrda <- dbrda(total ~ ., as.data.frame(env)) #perform dbRDA
ordiplot(pond.dbrda)
# first model only intercept
pond.dbrda.mod0 <- dbrda(total ~ 1, as.data.frame(env)) 
ordiplot(pond.dbrda.mod0) # no vectors constrained here, so its a simple MDS (PCoA)
# now model with all explanatory variables
pond.dbrda.mod1 <- dbrda(total ~ ., as.data.frame(env))
# now iterate through all combos of explanatory variables, return model with lowest AIC
# pond.dbrda <- ordiR2step(pond.dbrda.mod0,pond.dbrda.mod1, permutations = how(nperm = 200))
# couldnt get ordiR2step to work, error that upper scope cannot be reached
pond.dbrda <- ordistep(pond.dbrda.mod0,pond.dbrda.mod1, permutations = how(nperm = 200))
# look at model that was selected
pond.dbrda$call
pond.dbrda$anova
ordiplot(pond.dbrda)
# permutation tests to evaluate significance
permutest(pond.dbrda, permutations=999)
envfit(pond.dbrda, env[,c()], perm=999)
#calculate explained variation
pond.dbrda.explainvar1 <- round(pond.dbrda$CCA$eig[1] / sum(c(pond.dbrda$CCA$eig, pond.dbrda$CA$eig)), 3) *100
pond.dbrda.explainvar2 <- round(pond.dbrda$CCA$eig[2] / sum(c(pond.dbrda$CCA$eig, pond.dbrda$CA$eig)), 3) *100

# now plot selected model, define plot parameters
pond.vectors <- scores(pond.dbrda,display="bp")
plot(scores(pond.dbrda, display="wa"), xlab=paste("dbRDA 1 (",pond.dbrda.explainvar1,"%)", sep=""), 
     ylab=paste("dbRDA 2 (", pond.dbrda.explainvar2, "%)", sep=""),
     pch=16, cex=2.0, type="n", cex.lab=1.5, cex.axis=1.2, axes=FALSE);
axis(side=1, labels=T, lwd.ticks=2, cex.axis=1.2,las=1);
axis(side=2, labels=T, lwd.ticks=2, cex.axis=1.2,las=1);
abline(h=0,v=0,lty=3);
box(lwd=2);
points(scores(pond.dbrda,display="wa"), pch=19,cex=3,bg="gray",col="gray");
text(scores(pond.dbrda,display="wa"),labels=row.names(scores(pond.dbrda,display="wa")));
arrows(0,0,pond.vectors[,1], pond.vectors[,2], lwd=2, lty=1, length=0.2, col="red");
text(pond.vectors[,1], pond.vectors[,2], pos=3,labels=row.names(pond.vectors));
axis(side=3, lwd.ticks=2, cex.axis=1.2,las=1, col="red", lwd=2.2,
     at=pretty(range(pond.vectors[,1]))*2, labels=pretty(range(pond.vectors[,1])));
axis(side=4, lwd.ticks=2, cex.axis=1.2,las=1, col="red", lwd=2.2,
     at=pretty(range(pond.vectors[,2]))*2, labels=pretty(range(pond.vectors[,2])))

## plot output is odd because pond.vectors are values < 1 so they look tiny. is this an error in the code?

## note for Erica to think about -- way to add spec scores like from last week, with a cut off only showing the most correlated OTUs?
```

> ***Synthesis Answer***: According to the Mantel test, 7% of the variation in pond bacterial communities can be explained by association with the ponds' corresponding environmental variables. This is much lower than what was found with the Doubs river data, but the bacterial OTUs are much more diverse than the fish species in the Doubs. The constrained ordination via dbRDA indicated that the variable "chla" was associated with much of the variation in pond data.