---
title: "QB2023_Team_Project: Indiana Pond Bacterial Communities"
author: "Erica Nadolski and Joy O'Brien, Z620: Quantitative Biodiversity, Indiana University"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
geometry: margin = 2.54cm
editor_options: 
  chunk_output_type: console
---
### environment setup 
```{r}
package.list <- c("picante",'ape', 'seqinr', "vegan","fossil","reshape",
                  "devtools","BiocManager","ineq","labdsv",
                  "matrixStats","pROC",'phylobase', 'adephylo', 'geiger',
                  'stats', 'RColorBrewer', 'caper', 'phylolm', 'pmc', 
                  'ggplot2', 'tidyr', 'dplyr', 'phangorn', 'pander', 
                  'phylogram', 'dendextend', "tidyverse", "ggfortify")
for (package in package.list) {
  if (!require(package, character.only=TRUE, quietly=TRUE)) {
    install.packages(package)
    library(package, character.only=TRUE)
  }
}
```

#### Input and wrangle data
```{r, echo=FALSE}
# data 
Ponds.env <- as.data.frame(read.csv("/Users/ericanadolski/GitHub/team2/Ponds.env.csv", row.names=1, header = T, sep = ",", dec = ".")) 
Raw.count <- as.matrix(read.csv("/Users/ericanadolski/GitHub/team2/Ponds97.csv",   row.names=1, header = T, sep = ",", dec = "."))

# convert raw counts to relative counts 
Rel.count <- as.data.frame(decostand(Raw.count, method = "total"))

# most abundant species
spscore  <- sort(c(colSums(Rel.count)))
abundant <- as.data.frame(tail(spscore, 20))
selected <- rownames(abundant)

# add a source genetic material column (active/total communities)
# also add a site column to easily merge with environmental data 
Rel.count$site    <- gsub("-.*", "" , rownames(Raw.count))
Rel.count$source  <- gsub(".*-", "" , rownames(Raw.count))

# add a site column to environmental data 
Ponds.env$site    <- rownames(Ponds.env)
  
# collect all data in one table so everything matches 
allData <- left_join(Rel.count, Ponds.env, by = "site")

# for downstream analysis we need species matrix, factor columns (location and source community)
species_mat <- allData[,1:34059] # heatmap, PCoA
meta_data   <- allData[,34060:34081]
env_data    <- meta_data[,4:22]  

# standardize environmental data 
env_scaled  <- as.data.frame(scale(env_data, center = T))
```


### PCA of Environmental Variables to visualize variation

```{r pressure, echo=FALSE}
env.pca = prcomp(na.omit(meta_data[,4:22]), scale. = TRUE)
autoplot(env.pca, data = na.omit(meta_data), colour = 'Location')+
  theme_bw()
```

> ***Dataset Note***: We are using a Lennon lab dataset of microbial DNA and cDNA extracted from local ponds. There are 58 pond sites, and 34059 species (OTUs) across all the sites. There is high variance in abundance and evenness across all of the sites; based on exploratory rank abundance curves, there is high abundance of a few OTUs and a long tail of low-abundance OTUs.
```

### Alpha Diversity 
```{r}
# Rank Abundance Curve

RAC <- function(x="")+{
  x.ab = x[x > 0]
  x.ab.ranked = x.ab[order(x.ab, decreasing=TRUE)]
  as.data.frame(lapply(x.ab.ranked, unlist))
  return(x.ab.ranked)
}

# run RAC function
ponds.rac <- as.numeric(RAC(species_mat[1, ]))
length(ponds.rac)
max(ponds.rac)
min(ponds.rac)
plot.new()
pond.ranks <- as.vector(seq(1, length(ponds.rac)))
opar <- par(no.readonly = TRUE)
par(mar = c(5.1, 5.1, 4.1, 2.1))
plot(pond.ranks, log(ponds.rac), type = "p", axes =F,
     xlab = "Rank in abundance", ylab = "Abundance", 
     las = 1, cex.lab = 1.4, cex.axis = 1.25);
box();
axis(side = 1, labels = T, cex.axis = 1.25);
axis(side = 2, las = 1, cex.axis = 1.25, labels = c(1, 10, 100, 1000, 10000), at = log(c(1, 10, 100, 1000, 10000)))


## create new dataframe 
alpha_metrics <- data.frame(meta_data[,1:3])

## species richness 
alpha_metrics$richness <- specnumber(species_mat)

## Shannon's diversity
alpha_metrics$ShanH <- vegan::diversity(species_mat, index="shannon")

## inverse simpsons
alpha_metrics$invS <- vegan::diversity(species_mat, index="invsimpson")

ggplot(alpha_metrics, aes(x=source, y=richness, fill=Location))+ 
  geom_boxplot()+
  labs(title="Indiana Pond Species Richness",x="Source Genetic Material", y = "Richness")+
  geom_dotplot(binaxis='y', stackdir='center', dotsize=0.5, position=position_dodge(0.75))+
  theme_classic()

ggplot(alpha_metrics, aes(x=source, y=ShanH, fill=Location))+ 
  geom_boxplot()+
  labs(title="Indiana Pond Shannon Diversity",x="Source Genetic Material", y = "Shannon's H")+
  geom_dotplot(binaxis='y', stackdir='center', dotsize=0.5, position=position_dodge(0.75))+
  theme_classic()

ggplot(alpha_metrics, aes(x=source, y=invS, fill=Location))+ 
  geom_boxplot()+
  labs(title="Indiana Pond Inverse Simpson",x="Source Genetic Material", y = "Inverse Simpson")+
  geom_dotplot(binaxis='y', stackdir='center', dotsize=0.5, position=position_dodge(0.75))+
  theme_classic()

mean(alpha_metrics$richness)
mean(alpha_metrics$ShanH)
mean(alpha_metrics$invS)

```

### Beta Diversity - Visualization

```{r}
library(viridis)
# Bray Curtis resemblance matrix
total.db <- vegdist(species_mat, method="bray")

# Heatmap 
levelplot(as.matrix(total.db), aspect="iso", col.regions=inferno,
          xlab="Pond Site", ylab= "Pond Site", scales=list(cex=0.5), 
          main= "Bray-Curtis Distance")

# Wards cluster analysis
total.ward <- hclust(total.db, method= "ward.D2")

par(mar = c(1, 5, 2, 2) + 0.1)
plot(total.ward, main="Indiana Pond Bacteria: Ward's Clustering", ylab= "Squared Bray-Curtis Distance")

# Principal Component Analysis
total.pcoa <- cmdscale(total.db, eig=TRUE, k=3)

exvar1 <- round(total.pcoa$eig[1] / sum(total.pcoa$eig), 3) * 100
exvar2 <- round(total.pcoa$eig[2] / sum(total.pcoa$eig), 3) * 100
exvar3 <- round(total.pcoa$eig[3] / sum(total.pcoa$eig), 3) * 100
total.sum.eig <- sum(exvar1, exvar2, exvar3)

# PCoA Plot PC1 x PC2 colored by state park
par(mar = c(2, 2, 2, 2) + 3)
plot(total.pcoa$points[ ,1], total.pcoa$points[ ,2],
     #xlim = c(-0.55, -0.3),
     #ylim = c(-0.03, 0.01),
     xlab= paste("PCoA 1 (", exvar1, "%)", sep = ""),
     ylab= paste("PCoA 2 (", exvar2, "%)", sep = ""),
     pch = 16, cex = 2.0, type = "n", cex.lab = 1.5,
     cex.axis=1.2, axes=FALSE);
axis(side = 1, labels = T, lwd.ticks = 2, cex.axis = 1.2, las = 1);
axis(side = 2, labels = T, lwd.ticks = 2, cex.axis = 1.2, las = 1);
abline(h = 0, v = 0, lty = 3);
box(lwd = 2);                                        
points(total.pcoa$points[1:18,1], total.pcoa$points[1:18,2],
       pch = 20, cex = 2, bg = "red", col = "red"); # BC
points(total.pcoa$points[19:41,1], total.pcoa$points[19:41,2],
       pch = 20, cex = 2, bg = "green", col = "green"); # H
points(total.pcoa$points[42:58,1], total.pcoa$points[42:58,2],
       pch = 20, cex = 2, bg = "blue", col = "blue");# YW
legend(x="topleft",,legend=c("Brown County", "Hoosier Natl. Forest", "Yellowood"),
       fill=c("red","green","blue"))

```


### Beta Diversity - Hypothesis Testing 
### Mantel Test
```{r}
# Mantel test to test the hypothesis that pond assemblages are correlated with pond environmental variables.
# define matrices
# input right matrices 
OTU.dist <- vegdist(species_mat, method="bray")
pond.env.dist <- vegdist(env_data, method="euclid", na.rm = TRUE)

# mantel test
mantel(OTU.dist,pond.env.dist)
```

### hypothesis testing - nonmetric multidimensional scaling NMDS
```{r}
#NMDS ordination 
set.seed(123456) #set seed is to fix the random number generator so you get the same results 
#each time when you run the tests 
nmds <- metaMDS(species_mat, distance = "bray") 

# extract data from the ordination to plot 
# site scores
data.scores          <- as.data.frame(scores(nmds)$sites)  
data.scores$site     <- allData$site   
data.scores$source   <- allData$source
data.scores$location <- allData$Location 

# species scores technically we can also test species significance if we want to 
species.scores         <- as.data.frame(scores(nmds, "species"))  
species.scores$species <- rownames(species.scores)  

# filter species ordination scores with the most abundant species 
species.scores.abundant <- filter(species.scores, species %in% selected) 

# shortening names
species.scores.abundant$name    <- gsub("000*", "" , rownames(species.scores.abundant))

# permutation tests with factors (RNA/DNA and location )
distance  <- vegdist(species_mat, "bray")
set.seed(42)
permanova <- adonis2(distance ~ . , data = meta_data[2:3], permutations = 999, na.action = na.omit)
permanova # both are significant 

# environmental vectors (envfit only works with continuous )
set.seed(55)
fit        <- envfit(nmds, env_scaled, na.rm = TRUE, permutations = 999)

# vector data of enviromental fit
arrows     <- data.frame(fit$vector$arrows, R = fit$vectors$r, P = fit$vectors$pvals)
arrows$env <- rownames(arrows)
arrows.p   <- arrows[arrows$P < 0.05,] #select the significant variables 
arrows.p

# plot 
ggplot()+ 
  geom_hline(yintercept = 0, linetype = "dashed")+
  geom_vline(xintercept = 0, linetype = "dashed")+
  geom_point(data = data.scores, mapping = aes(x = NMDS1, y = NMDS2, colour = location, shape = source), size=3, alpha = .6) +
  theme_bw()+
  coord_equal()+
  geom_segment(data = arrows.p, aes(x = 0, y = 0, xend = NMDS1, yend = NMDS2), 
               arrow = arrow(length = unit(.2, "cm")*arrows.p$R),  color = "grey")+
  geom_text(data = arrows.p, aes(x = NMDS1, y = NMDS2, label = env), size=5)+
  geom_text(data = species.scores.abundant, aes(x = NMDS1, y = NMDS2, label = name), size = 3)# add abundant species scores

```
> ***Ordination Description***: It appears that the BC and HNF sites cluster more closely together into one cluster, and the YSF sites cluster more tightly into another cluster. According to the Mantel test, 23% of the variation in pond community similarity varies in correspondence with pond environmental similarity. The NMDS with envfit also shows that a variety of environmental variables covary with community composition along the first two NMDS axes, as well as a number of indicator species.